Loyalty's Limits: The Impact of Postmodernism and Political Polarization on Military Professionalism  (Prospectus)

### 2020-10-01 code/prospectusv1.Rmd
commit

---
title: "The Limits of Loyalty: Postmodernism, Polarization, and Military Professionalism  (Prospectus)"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |

  _This dissertation seeks to explore the relationship of postmodernism and political polarization and their impact on the US military profession. The fundamental question that I seek to answer is, "how is rising political polarization shaping the military as a profession in the United States, and what are the implications of these changes, both in terms of national security and the future of the military profession?" In answering this question, I develop a theory that, at its core, argues that the nature of rising political polarization fundamentally challenges the conception of loyalty among members of the military profession. I further argue that the strain of loyalty is especially problematic for military, as it is a profession that by its nature holds multiple loyalties simultaneously: to the Constitution, to the President as Commander-in-Chief, to Congress, and to the American People. Balancing these competing loyalties becomes more difficult, I argue, as political polarization worsens, because of the fundamental values - not mere policy differences - that are at the root of many polarized political battles today. In the dissertation, I seek to employ a mixed-methods approach that incorporates several case studies, as well as quantitative analysis, to support my theory. This dissertation will contribute to the fields of American politics, international security, and civil-military relations by further illuminating the mechanisms through which political polarization is impacting the military institution. Specifically, it will provide a contrarian view to recent civil-military relations scholars who have argued that the traditional concept of military professionalism should be significantly redesigned, if not abandoned entirely. This dissertation will argue the opposite: that what is needed most in politically polarizing times, such as those the US is in right now, is a return to and embrace of traditional conceptions of military professionalism - an effort that will require deliberate action by not just military leaders, but elected civilian leaders as well._   
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: researchpaper-ps904.bib # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs

@@

The question of if and how political polarization drives changes in the professional identity of the military spans several related literatures, most notable of which are American Politics and International Relations. Broadly speaking, this project is focused on how one aspect of domestic politics, political polarization, impacts a specific but important aspect of international relations, civil-military relations, which itself spans a wide and diverse literature. Because I intend to focus my work on the military as an institution, my dissertation will also rely on literature from the fields of sociology, organizational psychology, and ethics. I will frame my research question within each of these fields briefly before addressing why this question matters and should be studied in depth.

Within the American Politics literature, there is broad recognition that political polarization levels in contemporary America are higher now than at any point since the end of the American Civil War [@hare_polarization_2014], but there are multiple explanations for what has caused this trend, ranging from a growing divide over matters pertaining to race and religion [@abramowitz_great_2018], the size and role of government [@sides_two_2015], and disagreements over specific policy issues [@mann_its_2012]. Despite the herculean task of disentangling the roots of the causes of polarization, scholars agree that political polarization is a problem of sorts.

One of the major concerns regarding political polarization is its on impact on American foreign policy in a broad sense, and specifically, on America's ability to lead and impact the world. Kenneth Schultz, for instance, underscored four challenges that have emerged as a result of America's polarized domestic politics, including the ability to obtain bipartisan support for "risky undertakings," the ability to learn the "lessons of foreign policy failure," the inability to "make long term commitments to allies and adversaries" as a result of "dramatic policy swings from one administration to another," and the "vulnerability" of America's electoral system to foreign third party adversaries [@schultz_perils_2017]. Because the military is a central actor in shaping America's foreign policy, obtaining a better understanding of how political polarization shapes various military institutions specifically will yield findings that may bear on helping policymakers and academics help solve each of the problems Schultz identifies.

@@

Thankfully, a small but growing number of civil-military relations scholars have looked at aspects of this question in recent years. But the focus of these works has been, I believe, not on the impact of polarization on the military itself but on other important but peripheral components of the military's relationship with American society. These scholars have made important findings that speak to the state of current American civil-military relations. One recent dissertation completed by Michael Robinson at Stanford University in 2018, for instance, found that partisan affiliation significantly affects public credibility and trust in the military, perhaps resulting in or as a consequence of media "echo chambers" which may incentivize military leaders to offer opinions partisan opinions without fear of sanction [@robinson_danger_2018]. Robinson's work builds off of previous work by Heidi Urben in which she found that active military officers believe less strongly that retired military officers should remain apolitical or non-partisan in retirement [@urben_party_2013], a norm that has existed for decades or more. Together, Robinson and Urben join other scholars in providing a wealth of information regarding public opinion and the military, and in particular, how aspects of political polarization shape partisan differences regarding topics such as the use of force [@golby_jim_elite_2017] and military elite endorsement of political candidates [@golby_military_2012].

In their own right, these scholarly developments are interesting and important, but another debate in the civil-military relations field has recently emerged, or to be more correct, reawakened. At the center of this debate is the degree to which the concept of _military professionalism_ should be altered, if not discarded altogether [@brooks_paradoxes_2020]. The central question in this debate is whether the concept of _military professionalism_ actually accomplishes its intended goals: to simultaneously achieve military security from external threats, and to maintain civilian control of the military -- an argument developed in Samuel Huntington's 1957 work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_.

In this seminal work, which often serves as a point of departure for many discussions and debates within the field of civil-military relations, Huntington asserts that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced, and the second was a "societal imperative," which Huntington claims actually constrains the military in that the military can never become so alienated from society such that the military no longer represents the liberal democracy it defends [@huntington_soldier_1957, 1-3]. The societal imperative, Huntington claimed, was a further function of two distinct features of the United States. The first was America's liberal ideology [need to confirm this name of ideology], which Huntington saw as fundamentally suspicious of standing military forces, dating back to the Revolutionary War period [need cite]. Moreover, American liberal ideology for Huntington also values the individual extensively. The second feature of the societal imperative, claimed Huntington, was the US constitution, which clearly states that the military will always be controlled by civilians. Writing in the context of the seemingly long and dangerous Cold War, Huntington's point is that America needs a strong military that can win on the battlefield, but cautions that such a military cannot fundamentally alienate the character of American liberal society.

Figure one, below, shows a visual diagram of Huntington's logic in a political science framework (though as @feaver_civil-military_1996 points out, Huntington did not use such language when articulating his concepts). According to Huntington, the ultimate goal, or dependent variable, of civil-military relations in the American context is achieving both military security and maintaining civilian supremacy over the military. Huntington viewed the key independent variable that impacts and enables these goals while also resolving the tension generated by the opposing functional and societal imperatives, as described above, as adoption of a model of "objective civilian control" of civil-military relations, whereby political leaders maximize the professional autonomy of the military by establishing the ends and goals of policy, but allow military leaders to develop and implement the military plans to accomplish these goals [@huntington_soldier_1957, 80-85].

In making his argument, Huntington went to great lengths to describe an alternate world in which the pattern of civilian control, the key independent variable impacting military security and civilian supremacy of the military, was not "objective control," but "subjective control" [need cite for Huntington]. In this conception, Huntington saw civilian leaders making the military a political object through a variety of means, including the selection of military leaders sympathetic to partisan political causes [need cite], to the point that there was no longer any distinction between the military and civilian leadership [need cite]. To Huntington, adopting this form of control, and thus erasing what he saw as a necessary distinction between both the military and civilian leadership and society, would endanger American security. By virtue of its distinct mission and functional goals, Huntington asserted that the military needed to operate apart from partisan politics - something it could not do if a pattern of subjective control were to be adopted and implemented - in order to maintain military security.

@@

huntington_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("MSCS"="Military Security and\n Civilian Supremacy", 
                     "CC"="Patterns of\n Civilian Control\n 'Objective Control'", 
                     "FI"="Functional\n Imperative", 
                     "SI"="Societal\n Imperative",
                     "I"="Liberal Ideology",
                     "S"="Constitutional Structure")) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag() + labs(title="Figure One: Samuel Huntington's Theory\n Of Civilan Control of the Military", subtitle="As espoused in The Soldier and The State, 1957")

@@

Numerous critiques surrounding the Huntingtonian ideal of _military professionalism_ have been levied over the years. Most of these critiques, in one way or another, charge that a strict theoretical and practical separation between military and political operating spheres is fundamentally flawed. Many, if not all, of these critiques cite Clausewitz's famous dictum that war is a "mere continuation of policy by other means" [@clausewitz_war_1976, 69, 605], charging that war by its very nature requires the highest integration - not separation - of military and political activities. In essence, these critiques argue that the strict separation of the responsibilities between the military and civilians leads to poor strategic outcomes [@rapp_civil-military_2015; @brooks_paradoxes_2020].  Moreover, these criticisms can and do point largely to American failures in Iraq and Afghanistan to secure, at least as of yet, a more-lasting peace - failures that are ultimately, according to one scholar and former general, moral in nature, and shared between both civilians and the military [@dubik_just_2016].

As an extension of the critiques of Huntington, there has also been renewed focus and attention among civil-military relations scholars on numerous professional normative standards of the military. One of the most important yet difficult debates is an attempt by scholars to define where appropriate political behavior ends and partisan activity begins [@owens_military_2015; @golby_beyond_2015]. In the Huntingtonian framework, this is essentially the opposite side of the debate over the same coin in attempting to define for civilians what constitutes meddling in military affairs. One of the reasons this debate has become even more difficult, I believe, is because as political polarization increases, the lines between partisanship and politics, as well as politics and everything not politics, have blurred, as so much of the literature on American politics and polarization affirms.

@@

We also know that the military services in the United States place a premium on loyalty as a trait and concept. Indeed, of the different military services, the US Army in particular has extensively developed, in a normative sense, a framework entitled "The Army Ethic," in which loyalty constitutes a significant part of the Army profession [@us_department_of_the_army_adp_2019]. This ethic, which will be described in detail in later parts of this dissertation, is formally laid out and described in the Army's doctrinal publication entitled, _Army Leadership and the Profession, ADP 6-22_, and holds that as "stewards of the Army Profession" and as "committed professionals," Army leaders "embrace and uphold the Army Values and standards of the profession" [@us_department_of_the_army_adp_2019]. There are a total of seven "Army Values," which together spell the acronym, l-d-r-s-h-i-p, or "leadership," and the first of these values, loyalty, is followed by six others: duty, respect, selfless service, honor, integrity, and personal courage [@us_department_of_the_army_adp_2019, p. 1-9.].  The US Army states in its doctrinal manual on Army Leadership and the Profession, ADP 6-22, that the Army Values "embody the practical application of the Army Ethic. They encompass the enduring moral principles, beliefs, and laws that guide Army professionals in accomplishing the mission as well as their conduct in all aspects of life" [need cite here for ADP 6-22, p. 1-12]. The manual defines loyalty as, "loyalty: bear true faith and allegiance to the Constitution of the United States, the Army, your unit and other Soldiers" [need cite here for ADP 6-22.]

@@

# The Analytical Framework - Initial Arguments and Hypotheses

Given all that the existing scholarship says about the topics of political polarization, military professionalism, and concepts of loyalty and identity, the goal in my dissertation is to examine how these concepts are inter-related, and specifically, if and how polarization drives changes in the way that these concepts are inter-related. Further, I want to determine if these changes impact national security, and if so, in what ways?

Figure two below, shows an initial way to link some of the variables that I am interested in exploring throughout my dissertation. From left to right, this conceptual model suggests that aspects of political polarization are simultaneously driving changes in both the institutional credibility of the military (how society views the military), as well as changes in the conceptions of loyalty and identity (within the military). These two variables are, in turn, driving changes in the concept of military professionalism, my main dependent variable, within the military, and I suspect that these changes may, if not yet, potentially impact various military outcomes. By various military outcomes, I mean the very same military outcomes that other scholars do and have already explored, such as effectiveness on the battlefield, i.e winning wars and battles [@reiter_democracies_2002; @biddle_democracy_2004] and achieving strategic effectivness [@brooks_creating_2007; @brooks_shaping_2008]. While I intend to examine this system of variables, my focus will be on the link between increasing political polarization, the concepts of loyalty and identity, and military professionalism, both in order to focus my research but also because there has been more work done recently examining other variables, to include the impact of increasing political polarization on institutional credibility [@robinson_michael_danger_2018].

@@

problem_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("MO"="Changes in\n Military\n Outcomes", 
                     "MP"="Military\n Professionalism", 
                     "PP"="Increasing\n Political\n Polarization", 
                     "IC"="Institutional\n Credibility",
                     "LI"="Concepts of\n Loyalty and Identity")) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag() + labs(title="Figure One", subtitle="The Impact of Political Polarization on Military Professionalism")

@@

A primary argument is...

how increasing domestic political polarization impacts conceptions of loyalty in military professions, and to examine the implications of these changes in loyalty on broader security outcomes. For at least the past 75 years[need footnote here], the US military has prided itself on being not merely an organization charged with the defense of the nation, but a profession. That is, on being an institution marked by several unique traits, including: providing a unique service that requires special skills (the application and management of violence); establishing and maintaining standards of behavior and performance, as well as policing itself when members of that profession fail to uphold these standards; and building and maintaining trust between itself, the profession, with the client whom it serves (the American people) [insert source here].  But what happens if and when the objects of the military's multiple loyalties not merely disagree, but become severely divided ideologically? Can such an outcome exist, and if it can, is there empirical proof that such a trend is occurring? If so, what are the normative implications, both for the military as a profession, and for the nation's security, and are there actions that can be taken to halt such a trend?

## The Concept and Structure of Loyalty

In my dissertation, I situate changing conceptions of loyalty as a major sub component of the arrow running from political polarization to changes in military professionalism. At the same time, I do think that political polarization is impacting the credibility of the military institution, which in turn may likewise be driving changes in military professionalism. The specific mechanisms through which this is occurring, if at all, however, need to be developed, proposed, and tested.

At this stage in the dissertation phase, I do not know how polarization is impacting conceptions of loyalty, or even if it is, for sure. I strongly suspect that it is, however, for a couple of reasons. First, the military is, I believe, susceptible to charges of failing to uphold the ideal of loyalty based on the nature of the organization in that it ultimately is charged to enact violence on behalf of the state. Harold Lasswell said of military officers that their skill is the "management of violence" [need quote from Huntington]. As the bonds of loyalty forged on the battlefield are strong and different in ways than loyalty forged in other environments, I suspect that the military and its leaders will take very seriously any accusations that the corporate entity or individual officers are failing to be loyal to a number of ideals. In the Army, for example, the first "Army Value," as it is called by the military service,  which is summarized as the acronym, "LDRSHIP" (short for "leadership"), is the word "loyalty." [need cite for the Army values here].

A second reason that I suspect polarization is impacting the conception of loyalty among the military professions stems from some of the mechanisms that we know regarding how political polarization operates. Politics of American scholars, though not united in agreement regarding the causes of polarization, have agreed that polarization is working through at least three mechanisms, all of which, I believe, impact the military in its normative role qua non-partisan and apolitical American institution. First, there is a sharp increase in the number and scope of issues that have been subsumed by partisan debate and disagreement (a phenomenon referred to among scholars of American Politics as "conflict extension" [@layman_party_2002]). A second mechanism exists through the increase in the strength of negative feelings held by one partisan group towards those of the other side, a phenomenon referred to as "negative partisanship" [@abramowitz_great_2018]. Third, there is an increase in the salience of debates over issues that individual partisans consider core to their individual identities [@white_values_2002; @mann_its_2012; @sides_identity_2018]. Each of these three features of the politically polarized landscape strain the concept of military professionalism in important ways, and taken together, comprise a significant challenge to the individual military professional and thus, to the professional nature of the military as a whole. These challenges are theorized in Table 1, below.

>>>>

---
title: "Loyalty's Limits: The Impact of Postmodernism and Political Polarization on Military Professionalism  (Prospectus)"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |

  _This dissertation seeks to explore the impact of increasing political polarization on the professional identity of military officers and the military institution. I develop a theory that argues that the nature of rising political polarization fundamentally challenges the conception of loyalty among members of the military profession, and that such a strain is problematic for an institution such as the military that holds the virtue of loyalty in high regard, and by statutes possesses multiple, simultaneous loyalties: to the Constitution of the United States, to the President qua Commander-in-Chief, to Congress, to the American People, to one's fellow service members, and to one's self. Balancing these competing loyalties becomes more difficult as political polarization worsens, I argue, because of the fundamental and core values - not mere policy differences - that are at the root of many contemporary polarized political battles. Employing a mixed-methods approach that incorporates quantitative analysis and several historical case studies, this dissertation argues that the challenges to conceptions of loyalty among members of the military institution brought about by political polarization have the potential to impact normative concepts of military professionalism, which could be problematic not only for the character and identity of the military as an institution, but for American national security at large._   
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: researchpaper-ps904.bib # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs

@@

The question of if and how political polarization drives changes in the professional identity of the military spans several related literatures, most notable of which are American Politics and International Relations. Broadly speaking, this project is focused on how one aspect of domestic politics, political polarization, impacts a specific but important aspect of international relations, civil-military relations, which itself spans a wide and diverse literature. Because I intend to focus my work on the military as an institution, my dissertation will also rely on literature from the fields of sociology, organizational psychology, and ethics. I will frame my research question within each of these fields briefly, and to define important concepts as required, before addressing why this question matters and should be studied in depth.

Within the American Politics literature, there is broad recognition that political polarization levels in contemporary America are higher now than at any point since the end of the American Civil War [@hare_polarization_2014], but there are multiple explanations for what has caused this trend, ranging from a growing divide over matters pertaining to race and religion [@abramowitz_great_2018], the size and role of government [@sides_two_2015], and disagreements over specific policy issues [@mann_its_2012]. Despite the wide-ranging scholarship exploring the many facets of political polarization, virtually all scholars agree that contemporary levels of political polarization in the United States are historically abnormal and pose myriad problems.

One of the major concerns regarding political polarization is its on impact on American foreign policy in a broad sense, and specifically, on America's ability to lead and impact the world. Kenneth Schultz, for instance, underscored four challenges that have emerged as a result of America's polarized domestic politics, including the ability to obtain bipartisan support for "risky undertakings," the ability to learn the "lessons of foreign policy failure," the inability to "make long term commitments to allies and adversaries" as a result of "dramatic policy swings from one administration to another," and the "vulnerability" of America's electoral system to foreign third party adversaries [@schultz_perils_2017]. Because the military is a central actor in shaping America's foreign policy, obtaining a better understanding of how political polarization shapes various military institutions will yield findings that may help policymakers and academics help solve the problems Schultz identifies.

@@

Civil-military relations comprises an immense area of scholarship and interest, and deals with a correspondingly large number of questions and debates that have occupied the minds of political scientists, theorists, historians, ethicists, and sociologists for hundreds if not thousands of years. At its root, the study of civil-military relations examines what Peter Feaver calls the "problematique," or how to ensure that the military is strong enough to defeat threats to the polity without threatening the polity itself [@feaver_civil-military_1996; @feaver_civil-military_1999]. Feaver describes his "problematique" as "vexing" and a great "paradox" in that the military must be "strong enough to prevail in war," yet submissive to authority such that it cannot and will not "destroy or prey on the society it is intended to protect" [@feaver_civil-military_1999, p. 214].

Civil-military relations involves a wide angle of significant and important theoretical questions that involve the pull and push of power between institutions, the delineation of authority, and a host of other massive and consequential issues impacting how militaries perform in war, their relationship to the societies they serve, and other important questions. In the context of the US, which has never experienced or come close to a coup (thankfully), the study of civil-military relations often focuses on the relationship between military and government elites, or what Risa Brooks calls "leadership at the state's apex" [@brooks_risa_integrating_2019, p. 380]. Indeed, it is at this level that vital decisions are made, including to go to war, how to prosecute war, how to achieve and implement effective strategy, how to prepare for the next conflict, and how large the military and its budget should be. Each of these questions, and many, many more, involve questions regarding the proper role of the military and its leaders relative to both society and, perhaps more important, the leaders of government. A number of scholars have examined this question on normative grounds, but none has been more lasting and consequential than the late Samuel Huntington, whose 1957 seminal work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, often serves as the point of departure for many scholarly works on civil-military relations. Although there have been many critiques of Huntington over the years [for example, see @janowitz_professional_1960; @finer_man_1962; @cohen_supreme_2003; @feaver_armed_2003; @brooks_paradoxes_2020], it is absolutely essential to understand Huntington's concept of normative civil-military relations, and what lies at the heart of it - a theory of civilian control of the military that maximizes the concept of what Huntington called "military professionalism" [@huntington_soldier_1957].

In _The Soldier and the State_, Huntington asserts that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced, and the second was a "societal imperative," which Huntington claims actually constrains the military in that the military can never become so alienated from society such that the military no longer represents the liberal democracy it defends [@huntington_soldier_1957, 1-3]. The societal imperative, Huntington claimed, was a further function of two distinct features of the United States. The first was an "ideological constant," which referred to America's political ideology of liberalism [@huntington_soldier_1957, 143-162]. Huntington argued that liberalism as a political ideology stressed "individualism" [@huntington_soldier_1957, 90], and that American society was, since its founding, suspicious of standing military organizations and professional military forces. The second feature of the societal imperative, however, was the US Constitution, which Huntington claimed to be "conservative," and written by "men who feared concentrated political power and who provided for the widespread dispersion of that power among numerous governmental units" [@huntington_soldier_1957, 143]. The tension raised by these two imperatives - one functional stemming from external threat, the other societal stemming from the character of American political ideology and structure - needed to be solved, claimed Huntington. Writing in the context of the seemingly long and dangerous Cold War, Huntington's point is that America needs strong military that can win on the battlefield, but cautions that such a military cannot fundamentally alienate the character of American liberal society. Both goals - military security and civilian control - had to be obtained.

Figure 1, below, shows a visual diagram of Huntington's logic in a political science framework [@feaver_civil-military_1996]. As Peter Feaver points out in a 1996 article, Huntington did not use such language when developing and explaining his concept. According to Huntington, the ultimate goal, or dependent variable, of civil-military relations in the American context is achieving both military security and maintaining civilian supremacy over the military. Huntington viewed the key independent variable that impacts and enables these goals while also resolving the tension generated by the opposing functional and societal imperatives, as described above, the adoption of a model of "objective civilian control" of civil-military relations, whereby political and military leaders "maximize military professionalism," which includes granting the military a high degree of professional autonomy [@huntington_soldier_1957, 80-85].

@@

huntington_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("MSCS"="Military Security and\n Civilian Supremacy", 
                     "CC"="Patterns of\n Civilian Control\n 'Objective Control'", 
                     "FI"="Functional\n Imperative", 
                     "SI"="Societal\n Imperative",
                     "I"="Liberal Ideology",
                     "S"="Constitutional Structure")) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag() + labs(title="Figure 1: Samuel Huntington's Theory\n Of Civilan Control of the Military", subtitle="As espoused in The Soldier and The State, 1957")

@@

Huntington's argument in support of "objective control" of the military stands in contra-distinction, as he explains, to an alternate model of "subjective control" in which civilian leaders limit the power of the military by treating it as any other civilian interest group [@huntington_soldier_1957, 80-83]. In fact, Huntington admits that the US Constitution actually envisioned this "subjective" scheme of civilian control of the military [@huntington_soldier_1957, 163-168]. What the framers did not see, nor could they, argues Huntington, was the rise of the military as a profession in response to ever-increasing and dangerous external threats. It was because of these pernicious threats, argues Huntington, which the US military had witnessed, confronted, and battled during the 19th and 20th centuries during World War One, World War Two, and now the Cold War, that the military profession had risen as a separate and distinct body that could focus on the demands of military service. Such a separation and distinctness, claims Huntington, would need to be embraced by civilians and the militarym going forward into the future. In so doing, claims Huntington, is the only way that both military security and civilian control of the military - two often-competing imperatives - will be safeguarded.

Numerous critiques surrounding the Huntingtonian ideal of _military professionalism_ have been levied over the years, as Huntington did not attempt to precisely define it. Most of these critiques, in one way or another, charge that a strict theoretical and practical separation between military and political operating spheres is fundamentally flawed. Many, if not all, of these critiques cite Clausewitz's famous dictum that war is a "mere continuation of policy by other means" [@clausewitz_war_1976, 69, 605], charging that war by its very nature requires the highest integration - not separation - of military and political activities. In essence, these critiques argue that the strict separation of the responsibilities between the military and civilians leads to poor strategic outcomes [@rapp_civil-military_2015; @brooks_paradoxes_2020].  Moreover, these criticisms can and do point largely to American failures in Iraq and Afghanistan to secure, at least as of yet, a more-lasting peace - failures that are ultimately, according to one scholar and former general, moral in nature, and shared between both civilians and the military [@dubik_just_2016].

As an extension of the critiques of Huntington, there has also been renewed focus and attention among civil-military relations scholars on numerous professional normative standards of the military. One of the most important -  yet difficult - debates is an attempt by scholars to define where appropriate political behavior ends and partisan activity begins [@owens_military_2015; @golby_beyond_2015]. In the Huntingtonian framework, this is essentially the opposite side of the debate over the same coin in attempting to define for civilians what constitutes meddling in military affairs. One of the reasons this debate has become even more difficult, I believe, is because as political polarization increases, the lines between partisanship and politics, as well as politics and everything not politics, have blurred, as so much of the literature on American politics and polarization affirms.

@@

We also know that the military services in the United States place a premium on loyalty as a trait and concept. Indeed, of the different military services, the US Army in particular has extensively developed, in a normative sense, a framework entitled "The Army Ethic," in which loyalty constitutes a significant part of the Army profession [@us_department_of_the_army_adp_2019]. This ethic, which will be described in detail in later parts of this dissertation, is formally laid out and described in the Army's doctrinal publication entitled, _Army Leadership and the Profession, ADP 6-22_, and holds that as "stewards of the Army Profession" and as "committed professionals," Army leaders "embrace and uphold the Army Values and standards of the profession" [@us_department_of_the_army_adp_2019]. There are a total of seven "Army Values," which together spell the acronym, l-d-r-s-h-i-p, or "leadership," and the first of these values, loyalty, is followed by six others: duty, respect, selfless service, honor, integrity, and personal courage [@us_department_of_the_army_adp_2019, p. 1-9.].  The US Army states in its doctrinal manual on Army Leadership and the Profession, ADP 6-22, that the Army Values "embody the practical application of the Army Ethic. They encompass the enduring moral principles, beliefs, and laws that guide Army professionals in accomplishing the mission as well as their conduct in all aspects of life" [@us_department_of_the_army_adp_2019, p. 1-12]. The manual defines loyalty as, "loyalty: bear true faith and allegiance to the Constitution of the United States, the Army, your unit and other Soldiers" [@us_department_of_the_army_adp_2019]. The virtue of loyalty is likewise extolled in a host of codes and mantras that have played a central part of the American military since its inception. In the middle of the West Point Cadet Prayer is the eloquently written plea, "make us to choose the harder right instead of the easier wrong" [@wheat_west_1918]. Even Huntington spent a great deal talking about the importance of loyalty as a trait to the very essence of the military profession. He recognized that in order for civilian control to be effective, the military profession had to organize itself as a "hierarchy of obedience," which required "loyalty and obedience" to stand out as "the highest military virtues" [@huntington_soldier_1957, 73].

@@

Thankfully, a growing number of civil-military relations scholars have looked at various aspects of this question - the impact of the political polarization on civil-military relations - in recent years. But the focus of these works has been, I believe, not on the impact of polarization on the military itself but on other important but peripheral components of the military's relationship with American society. One recent dissertation completed by Michael Robinson at Stanford University in 2018, for instance, found that partisan affiliation significantly affects public credibility and trust in the military, perhaps resulting in or as a consequence of media "echo chambers" which may incentivize military leaders to offer opinions partisan opinions without fear of sanction [@robinson_danger_2018]. Robinson's work builds off of previous work by Heidi Urben in which she found that active military officers believe less strongly that retired military officers should remain apolitical or non-partisan in retirement [@urben_party_2013], a norm that has existed for decades or more. Together, Robinson and Urben join other scholars in providing a wealth of information regarding public opinion and the military, and in particular, how aspects of political polarization shape partisan differences regarding topics such as the use of force [@golby_jim_elite_2017] and military elite endorsement of political candidates [@golby_military_2012].

# The Analytical Framework - Initial Argument

## How Polarization Impacts Loyalty as a Concept

In this section of the prospectus, I will lay out an initial argument, as well as accompanying hypotheses, which I intend to explore and test in my dissertation. Figure 2, below, shows an initial conception of the linkages between several of the variables discussed in the previous section. From left to right, this conceptual model suggests that worsening political polarization is driving changes in both the institutional credibility of the military (how society views the military) and changes in the conceptions of loyalty and identity (within the military). These two independent variables are, I suggest, driving subsequently driving changes in the concept of military professionalism, the main dependent variable of analysis in my dissertation. Following Huntington's model, I also suggest that changes in military professionalism can impact a variety of outcomes, by which I mean the very same outcomes that other scholars do and have already explored, such as effectiveness on the battlefield [@reiter_democracies_2002; @biddle_democracy_2004] and achieving strategic effectivness [@brooks_creating_2007; @brooks_shaping_2008]. While I intend to examine the broad system of variables shown in Figure 2, the focus of my dissertation, as well as the least-explored linkage in this model, is the relationship between political polarization and changes in the concept of loyalty and identity, and how such changes manifest themselves in the variation of military professionalism. For emphasis and clarity, I do not think that the other link - that between political polarization and institutional credibility - is unimportant. Rather, this link seems to be the focus of more-recent scholarship [@robinson_michael_danger_2018].

@@

problem_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("MO"="Changes in\n Military\n Outcomes", 
                     "MP"="Military\n Professionalism", 
                     "PP"="Increasing\n Political\n Polarization", 
                     "IC"="Institutional\n Credibility",
                     "LI"="Concepts of\n Loyalty and Identity")) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag() + labs(title="Figure 2", subtitle="The Impact of Political Polarization on Military Professionalism")

@@

A primary contention from the system visualized above is that as political polarization in the domestic sphere increases, the military faces increased challenges in how both individual military leaders as well as the institution conceptualizes, adopts, and exercises loyalty, and by extension, professional identity. This expectation stems from several features that scholars of American politics have raised regarding the current politically polarized climate within the United States. First, scholars have noted that political polarization raises the number and scope of issues that are subsumed by partisan debate and disagreement (a phenomenon referred to among scholars of American Politics as "conflict extension" [@layman_party_2002]). A second feature of contemporary political polarization is the prominence of  negative feelings held by one partisan group towards another, a phenomenon referred to as "negative partisanship" [@abramowitz_great_2018]. A third feature is an increase in the salience of debates over issues that individual partisans consider core to their individual identities [@white_values_2002; @mann_its_2012; @sides_identity_2018]. I contend that these features impact individuals serving in the military, including military leaders, as they do members of the domestic public writ large, and that these impacts are problematic for members of an institution that normatively exists as an apolitical and non-partisan actor. Table 1, below, lays out how these three prominent features of domestic political polarilzation may impact the non-partisan and apolitical nature of the normative military professional and institution.

\singlespacing
\begin{table}[ht]
\begin{footnotesize}
\caption{How Features of Contemporary Political Polarization Impact Normative Military Professionalism} % title of Table
\centering % used for centering table
\begin{tabular}{|p{5cm}|p{5cm}|p{5cm}|p{5cm}|}
    \hline
    \textbf{Features of Increasing Political Polarization} & \textbf{Associated Mechanism} & \textbf{Potential Impact on Military Professionalism}  \\ \hline
    \textbf{Conflict Extension} & Less politically-neutral space exists as partisan disagreements subsume a greater number/scope of issues. & Military Professionals and the Institution find it increasingly difficult to "stay out" of partisan politics as the territory of what is considered partisan politics grows. \\ \hline
    \textbf{Negative Partisanship} & Strong feelings of personal dislike and disdain arise against those who hold to opposing political views. There is a belief that those on the other side of the political aisle are not just mistaken over policy, but greatly wrong and flawed as individuals. & Even among military leaders, there is a greater willingness to confront or take issue with those who hold opposing views.\\ \hline
    \textbf{Individuals believe that "core" issues are increasingly at stake} & The salience of one's identity and core values is raised as partisans believe essential issues are at stake in the political arena. Associated with this is a belief that not just policy preference, but moral right and wrong, are at stake. & Military professionals, who hold multiple institutional loyalties, face an increasingly difficult task of balancing these loyalties, especially as the objects of these loyalties are increasingly associated with or attached to notions of moral right and wrong. \\ \hline
\end{tabular}
\label{table:comparison} % is used to refer this table in the text
\end{footnotesize}
\end{table}
\doublespacing

The first way in which the apolitical and non-partisan ethos of the military is challenged as a result of increasing political polarization is that there is increasing difficulty for individuals to remain politically neutral, non-partisan, or otherwise apolitical - a hallmark of military professionalism. This happens because as "conflict extension" [@layman_party_2002] occurs, an increasing number of issues are brought into partisan battlegrounds, and thus, there is simply less politically-neutral territory available. Thus, an actor who wishes to remain non-partisan, such as the military, may be increasingly less able to do so, as a result of this feature.

A second challenge military leaders and the institution faces in a polarized climate is from the "negative partisanship"[@abramowitz_great_2018] that dominates the polarized American political landscape. As citizens hold increasingly strong and negative feelings towards those who side differently on a host of issues, there may be a tendency for members of the military institution to harbor similar feelings. Even if impact of this feature on the military and its leaders, due to training and professional development that emphasizes the importance of upholding apolitical and non-partisan behavioral norms, is somewhat muted, the point is that members of the military institution are likely experiencing the same phenomenon - even if it is to a lesser degree - as the country's citizens.

A third, and I contend the most impactful, challenge that normative military professionalism faces stems from the belief by individuals that "core" issues are at stake in the current polarized political climate. "Core" issues are difficult to define precisely [need cite], but the main point is that these issues often involve perceptions of moral right and wrong, and not just simple policy preferences. Moreover, some of the most polarizing issues are often associated with issues related to an individual's identity, such as race, gender, immigration [need cite]. The rise of "identity politics" in America as political polarization has worsened illustrates this point [need cite].
